~\Large Stickle!

### 2023-05-10 manuscript/paper.Rmd
Edited Paper

---
title: |
  |  \Large Stickle!
author: | 
  | \large Matthew Stuart \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mstuart@luc.edu`](mailto:mstuart@luc.edu) \vspace{-1mm}
  |
  | \large Akhil Ghosh \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`aghosh@luc.edu`](mailto:aghosh@luc.edu) \vspace{-1mm}
  |
  | \large Yoel Stuart \vspace{-1.1mm}
  | \normalsize Department of Biology \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`ystuart@luc.edu`](mailto:ystuart@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Center for Data Science and Consulting; Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Evewryone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

# Models  {#sec:models}
## Imputation model 
Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times p$ matrix of the $p$ phenotypes of interest of the fish.  Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.  In this setting, $\boldsymbol{W}_{obs}$ is a $n_{extant} \times 1$ vector and $\boldsymbol{W}_{obs}$ is a $n_{fossil} \times 1$ vector, perfectly corresponding to the observable gender of the modern extants and the unobservable gender of the fossil data, respectively.

@@

## OU model 
After imputing sex, we fit an OU model for different phenotypes of interest. We only look at $n_{fossil}$ fossil observations.  
WLOG we choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.

@@

$$
\begin{aligned}
X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
\end{aligned}
$$
We then fit the time and gender dependent means, $\mu_{ft}$ and $\mu_{mt}$, to an autoregressive process of lag 1, which is the discretization of an OU process.  For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for the mean structure defined as is

$$
\begin{aligned}
\mu_{gt} \iid \mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
\end{aligned}
$$
In this setting, $\theta_g, g \in \{f,m\}$ is the defined as the overall long-run mean of the phenotype of interest for each gender, and $\kappa$ represents how quickly over time $\mu_{gt}$ reverts back to $\theta_g$; the larger the value of $\kappa$, the slower $\mu_{gt}$ reverts back to $\theta_g$.

@@

$$
\begin{aligned}
\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
\end{aligned}
$$
This model choice is to preserve stationarity, i.e. $p(\mu_{gs}) = p(\mu_{gt})$ for $s,t \in \{1,\dots,T\}$.

Priors: 
$$
\sigma \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}}
$$
$$
\tau \sim \mathcal{N}(0,100)I_{\{\tau > 0\}}
$$

$$
\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
$$
$$
\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
$$

>>>>

---
title: |
  |  \Large Stickle!
author: | 
  | \large Matthew Stuart \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`mstuart@luc.edu`](mailto:mstuart@luc.edu) \vspace{-1.1mm}
  |
  | \large Akhil Ghosh \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`aghosh@luc.edu`](mailto:aghosh@luc.edu) \vspace{-1.1mm}
  |
  | \large Yoel Stuart \vspace{-1.1mm}
  | \normalsize Department of Biology \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`ystuart@luc.edu`](mailto:ystuart@luc.edu) \vspace{-1.1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Center for Data Science and Consulting; Department of Mathematics and Statistics \vspace{-1.1mm}
  | \normalsize Loyola University Chicago \vspace{-1.1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1.1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1.1mm}
abstract: |
  | Evewryone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

# Models  {#sec:models}
## Imputation model 
Let $\boldsymbol{W}$ be an $(n_{extant} + n_{fossil}) \times 1$ vector of the covariate gender of the stickleback fish and $\boldsymbol{X}$ be an $(n_{extant} + n_{fossil}) \times k$ matrix of the $k$ phenotypes of interest of the fish.  Because $\boldsymbol{W}$ contains missing data, we further define $\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T$ where $\boldsymbol{W}_{obs}$ and $\boldsymbol{W}_{mis}$ are the observed and missing data of the covariate gender, respectively.  In this setting, $\boldsymbol{W}_{obs}$ is a $n_{extant} \times 1$ vector and $\boldsymbol{W}_{obs}$ is a $n_{fossil} \times 1$ vector, perfectly corresponding to the observable gender of the modern extants and the unobservable gender of the fossil data, respectively.

@@

## OU model 
After imputing sex, we fit an OU model for different phenotypes of interest.  WLOG we choose one of the columns of $X$ to be the target phenotype and our analysis is repeated for all phenotypes of interest.

@@

$$
\begin{aligned}
X_{ti} \iid \left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
\end{aligned}
$$
In this setting, for $g \in \{f,m\}$, $\theta_g$, is the defined as the overall mean of the phenotype of interest for each gender, and $u_{gt}$ are the gender specific random effects, measuring the difference between the gender and time-specific mean and the overall gender-specific mean, which we fit to an autoregressive process of lag 1, which is the discretization of an OU process.  For $t = 2,\ldots,T$ and $g \in \{f,m\}$, the AR1 model for is defined as

$$
\begin{aligned}
u_{g,t} \iid \mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\end{aligned}
$$
$\kappa_g$ represents the gender-specific correlation between the previous and current time periods' random effects.  In addition, $\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h$ for $h \in \mathbb{N}$.

@@

$$
\begin{aligned}
u_{g,1} \iid \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\end{aligned}
$$
This model choice is to make the distribution of the random effects stationary, i.e. $p(\mu_{g,s}) = p(\mu_{g,t})$ for $s,t \in \{1,\dots,T\}$.

Priors: 
$$
\begin{aligned}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
\kappa & \iid \mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
\theta_g & \iid \mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\end{aligned}
$$

### 2023-05-10 manuscript/paper.tex
Edited Paper

\title{~\Large Stickle!}
\author{\large Matthew Stuart \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:mstuart@luc.edu}{\texttt{mstuart@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Akhil Ghosh \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:aghosh@luc.edu}{\texttt{aghosh@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Yoel Stuart \vspace{-1.1mm}\\
\normalsize Department of Biology \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:ystuart@luc.edu}{\texttt{ystuart@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Center for Data Science and Consulting; Department of
Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1mm}}
\date{}

@@

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish and
\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times p\) matrix
of the \(p\) phenotypes of interest of the fish. Because
\(\boldsymbol{W}\) contains missing data, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T\)
where \(\boldsymbol{W}_{obs}\) and \(\boldsymbol{W}_{mis}\) are the
observed and missing data of the covariate gender, respectively. In this
setting, \(\boldsymbol{W}_{obs}\) is a \(n_{extant} \times 1\) vector
and \(\boldsymbol{W}_{obs}\) is a \(n_{fossil} \times 1\) vector,
perfectly corresponding to the observable gender of the modern extants
and the unobservable gender of the fossil, respectively.

@@

After imputing sex, we fit an OU model for different phenotypes of
interest. We only look at \(n_{fossil}\) fossil observations.\\
WLOG we choose one of the columns of \(X\) to be the target phenotype
and our analysis is repeated for all phenotypes of interest.

@@

\[
\begin{aligned}
X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\mu_{ft},\sigma^2) & W_{ti} = \text{Female} \\ \mathcal{N}(\mu_{mt},\sigma^2) & W_{ti} = \text{Male} \end{array}\right.
\end{aligned}
\] We then fit the time and gender dependent means, \(\mu_{ft}\) and
\(\mu_{mt}\), to an autoregressive process of lag 1, which is the
discretization of an OU process. For \(t = 2,\ldots,T\) and
\(g \in \{f,m\}\), the AR1 model for the mean structure defined as is

\[
\begin{aligned}
\mu_{gt} \overset{iid}{\sim}\mathcal{N}(\kappa \mu_{g,t-1} + (1 - \kappa)\theta_g, \tau^2).
\end{aligned}
\] In this setting, \(\theta_g, g \in \{f,m\}\) is the defined as the
overall long-run mean of the phenotype of interest for each gender, and
\(\kappa\) represents how quickly over time \(\mu_{gt}\) reverts back to
\(\theta_g\); the larger the value of \(\kappa\), the slower
\(\mu_{gt}\) reverts back to \(\theta_g\).

@@

\[
\begin{aligned}
\mu_{g1} \sim \mathcal{N}\left(0,\frac{\tau^2}{1-\kappa^2}\right).
\end{aligned}
\] This model choice is to preserve stationarity,
i.e.~\(p(\mu_{gs}) = p(\mu_{gt})\) for \(s,t \in \{1,\dots,T\}\).

Priors: \[
\sigma \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}}
\] \[
\tau \sim \mathcal{N}(0,100)I_{\{\tau > 0\}}
\]

\[
\kappa \sim \mathcal{N}(0,25)I_{\{0 < \kappa < 1\}} \\
\] \[
\theta_g \sim \mathcal{N}(54,625)I_{\{\theta_g > 0\}} \\
\]

>>>>

\title{~\Large Stickle!}
\author{\large Matthew Stuart \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:mstuart@luc.edu}{\texttt{mstuart@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Akhil Ghosh \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:aghosh@luc.edu}{\texttt{aghosh@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Yoel Stuart \vspace{-1.1mm}\\
\normalsize Department of Biology \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:ystuart@luc.edu}{\texttt{ystuart@luc.edu}}
\vspace{-1.1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Center for Data Science and Consulting; Department of
Mathematics and Statistics \vspace{-1.1mm}\\
\normalsize Loyola University Chicago \vspace{-1.1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1.1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1.1mm}}
\date{}

@@

Let \(\boldsymbol{W}\) be an \((n_{extant} + n_{fossil}) \times 1\)
vector of the covariate gender of the stickleback fish and
\(\boldsymbol{X}\) be an \((n_{extant} + n_{fossil}) \times k\) matrix
of the \(k\) phenotypes of interest of the fish. Because
\(\boldsymbol{W}\) contains missing data, we further define
\(\boldsymbol{W} = (\boldsymbol{W}_{obs}^T,\boldsymbol{W}_{mis}^T)^T\)
where \(\boldsymbol{W}_{obs}\) and \(\boldsymbol{W}_{mis}\) are the
observed and missing data of the covariate gender, respectively. In this
setting, \(\boldsymbol{W}_{obs}\) is a \(n_{extant} \times 1\) vector
and \(\boldsymbol{W}_{obs}\) is a \(n_{fossil} \times 1\) vector,
perfectly corresponding to the observable gender of the modern extants
and the unobservable gender of the fossil data, respectively.

@@

After imputing sex, we fit an OU model for different phenotypes of
interest. WLOG we choose one of the columns of \(X\) to be the target
phenotype and our analysis is repeated for all phenotypes of interest.

@@

\[
\begin{aligned}
X_{ti} \overset{iid}{\sim}\left\{\begin{array}{ll} \mathcal{N}(\theta_f + u_{ft},\sigma^2), & W_{ti} = \text{Female} \\ \mathcal{N}(\theta_m + u_{mt},\sigma^2), & W_{ti} = \text{Male} \end{array}\right.
\end{aligned}
\] In this setting, for \(g \in \{f,m\}\), \(\theta_g\), is the defined
as the overall mean of the phenotype of interest for each gender, and
\(u_{gt}\) are the gender specific random effects, measuring the
difference between the gender and time-specific mean and the overall
gender-specific mean, which we fit to an autoregressive process of lag
1, which is the discretization of an OU process. For \(t = 2,\ldots,T\)
and \(g \in \{f,m\}\), the AR1 model for is defined as

\[
\begin{aligned}
u_{g,t} \overset{iid}{\sim}\mathcal{N}(\kappa_g u_{g,t-1} , \tau^2).
\end{aligned}
\] \(\kappa_g\) represents the gender-specific correlation between the
previous and current time periods' random effects. In addition,
\(\text{cor}(u_{g,t},u_{g,t+h}) = (\kappa_g)^h\) for
\(h \in \mathbb{N}\).

@@

\[
\begin{aligned}
u_{g,1} \overset{iid}{\sim}\mathcal{N}\left(0,\frac{\tau^2}{1-\kappa_g^2}\right).
\end{aligned}
\] This model choice is to make the distribution of the random effects
stationary, i.e.~\(p(\mu_{g,s}) = p(\mu_{g,t})\) for
\(s,t \in \{1,\dots,T\}\).

Priors: \[
\begin{aligned}
\sigma & \sim \mathcal{N}(0,100)I_{\{\sigma > 0\}} \\
\tau & \sim \mathcal{N}(0,100)I_{\{\tau > 0\}} \\
\kappa & \overset{iid}{\sim}\mathcal{N}(0.5,1)I_{\{-1 < \kappa_g < 1\}} \\
\theta_g & \overset{iid}{\sim}\mathcal{N}(54,400)I_{\{\theta_g > 0\}}
\end{aligned}
\]
